An introduction to wavelets using R

### 2022-02-06 text/14/wavelets.Rmd
Minor update to wavelets.Rmd
Reconstruction section of wavelets.Rmd

The calculation of the wavelet decomposition is an iterative process. First, you need the coefficients underlying the d4 wavelet. At the highest possible frequency, there are four coefficients.

@@

```{r low-pass}
h <- d4.rwt$h.1
h
```

@@

The V1 coefficients represent the running dot product of the signal with the g vector. You skip every other data point in the running dot product, and you rotate the dot product around to the end when reach the right hand side to avoid edge effects. The skipping past the second, fourth, sixth, etc. observation in the running dot product is described in some sources as down-sampling.

@@

$\begin{bmatrix} h_1 & h_2 & h_3 & h_4 &     &     & ... \\ &     & h_1 & h_2 & h_3 & h_4 & ... \\  . \\ . \\ . \\    &     &     &     &     &     & ... & h_1 & h_2 & h_3 & h_4 \\ h_3 & h_4 &     &     &     &     & ... &     &     & h_1 & h_2 \end{bmatrix}$

@@

c(W1[   1], -h %*% ecg[1:4])
c(W1[   2], -h %*% ecg[3:6])
c(W1[   3], -h %*% ecg[5:8])
c(W1[1023], -h %*% ecg[2045:2048])
c(W1[1024], -h %*% ecg[c(2047:2048, 1:2)])
```

@@

c(W2[  1], -h %*% V1[1:4])
c(W2[  2], -h %*% V1[3:6])
c(W2[  3], -h %*% V1[5:8])
c(W2[511], -h %*% V1[1021:1024])
c(W2[512], -h %*% V1[c(1023:1024, 1:2)])
```

@@

c(W3[  1], -h %*% V2[1:4])
c(W3[  2], -h %*% V2[3:6])
c(W3[  3], -h %*% V2[5:8])
c(W3[255], -h %*% V2[509:512])
c(W3[256], -h %*% V2[c(511:512, 1:2)])
```

@@

First you dilute the W and V values at the top of the pyramid.

```{r reconstuct V8}
V9_diluted <- c(V9[1], 0, V9[2], 0, V9[3], 0, V9[4], 0)
W9_diluted <- c(W9[1], 0, W9[2], 0, W9[3], 0, W9[4], 0)

@@

c(V8[1], rev(h) %*% V9_diluted[1:4] + rev(g) %*% W9_diluted[1:4])
c(V8[2], rev(h) %*% V9_diluted[2:5] + rev(g) %*% W9_diluted[2:5])
V9_coefficients <- rep(0, 2048)
V9_coefficients[1:8] <- wt.9$y[1:8]
V9_signal <- as.vector(midwt(V9_coefficients, d4.rwt$h.0, L=9)$x)
{
  graph_without_axis(V9_signal, "V9", limits=range(ecg))
  graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
}
```

W8d <- dilute(W8)
V8d <- dilute(V8)
 W8d, g, 14)+V8d,h,14),V7[1],"V7[1]")
V7[1]
By hand -3.315508
Using R -3.315508
-W8d, g, 15)-V8d,h,15),V7[2],"V7[2]")
V7[2]
By hand -2.398146
Using R -2.398146
 W8d, g, 16)+V8d,h,16),V7[3],"V7[3]")
V7[3]
By hand -0.2769286
Using R -0.2769286
-W8d, g,  1)-V8d,h, 1),V7[4],"V7[4]")
V7[4]
By hand 1.782748
Using R 1.782748
 W8d, g,  2)+V8d,h, 2),V7[5],"V7[5]")
V7[5]
By hand 5.25007
Using R 5.25007
-W8d, g,  3)-V8d,h, 3),V7[6],"V7[6]")
V7[6]
By hand 4.7496
Using R 4.7496

You dilute the V8 and W8 coefficients by inserting zeros between every element, compute a running dot product of both the g and the h vectors, and add the results together. For some reason, you have to multiply every other value by -1. I don't know why. It's one of those scaling things. Also notice that you start at almost the very end of V8. This is caused by edge effects.

Once you have V7, repeat the process on the diluted V7 and W7 vectors to get V6.

W7d <- dilute(W7)
V7d <- dilute(V7)
 W7d, g, 30)+V7d,h,30),V6[1],"V6[1]")
V6[1]
By hand -2.096562
Using R -2.096562
-W7d, g, 31)-V7d,h,31),V6[2],"V6[2]")
V6[2]
By hand -2.294492
Using R -2.294492
 W7d, g, 32)+V7d,h,32),V6[3],"V6[3]")
V6[3]
By hand -2.441797
Using R -2.441797
-W7d, g,  1)-V7d,h, 1),V6[4],"V6[4]")
V6[4]
By hand -1.265345
Using R -1.265345
 W7d, g,  2)+V7d,h, 2),V6[5],"V6[5]")
V6[5]
By hand -0.6650157
Using R -0.6650157
-W7d, g,  3)-V7d,h, 3),V6[6],"V6[6]")
V6[6]
By hand 0.08738041
Using R 0.08738041

And so forth.

W1d <- dilute(W1)
V1d <- dilute(V1)
 W1d, g, 2046)+V1d,h,2046),ecg[1],"ecg[1]")
ecg[1]
By hand -0.104773
Using R -0.104773
-W1d, g, 2047)-V1d,h,2047),ecg[2],"ecg[2]")
ecg[2]
By hand -0.093136
Using R -0.093136
 W1d, g, 2048)+V1d,h,2048),ecg[3],"ecg[3]")
ecg[3]
By hand -0.0815
Using R -0.0815
-W1d, g,  1)-V1d,h, 1),ecg[4],"ecg[4]")
ecg[4]
By hand -0.116409
Using R -0.116409
 W1d, g,  2)+V1d,h, 2),ecg[5],"ecg[5]")
ecg[5]
By hand -0.0815
Using R -0.0815
-W1d, g,  3)-V1d,h, 3),ecg[6],"ecg[6]")
ecg[6]
By hand -0.116409
Using R -0.116409

At the very end, the calculations on the diluted V1 and W1 values give you the original signal.

You can see what the wavelet coefficients at each level contribute by zeroing out some of the components and then inverting it.

V8 and W8
W7
W6
W5
W4
W3
W2
W1
ecg
Axis

Notice that the sharp spikes are captured in W1, W2, and W3. The slower overall undulation is captured by V8 and W8. Some of the intermediate squiggles between the sharp spikes are captured in the remaining wavelets.

>>>>

You don't actually use the larger wavelets shown above. Instead you apply the most extreme wavelets in an iterative method that reduces the amount of data by half at each iteration. This makes the processing fast and easy. The most extreme D4 wavelets have only four coefficients.

@@

Note: To get the math to work on my end, tThe low pass filter provided by the rwt package needs to be negated.

```{r low-pass}
h <- -d4.rwt$h.1
h
```

@@

The V1 coefficients represent the running dot product of the signal with the g vector. You skip every other data point in the running dot product (ignore the gray rows), and you rotate the dot product around to the end when reach the right hand side to avoid edge effects. The skipping past the second, fourth, sixth, etc. observation in the running dot product is described in some sources as down-sampling.

![Figure 1. G matrix](http://www.pmean.com/new-images/14/wavelets-01.png)

@@

![Figure 2. H matrix](http://www.pmean.com/new-images/14/wavelets-02.png)

@@

c(W1[   1], h %*% ecg[1:4])
c(W1[   2], h %*% ecg[3:6])
c(W1[   3], h %*% ecg[5:8])
c(W1[1023], h %*% ecg[2045:2048])
c(W1[1024], h %*% ecg[c(2047:2048, 1:2)])
```

@@

c(W2[  1], h %*% V1[1:4])
c(W2[  2], h %*% V1[3:6])
c(W2[  3], h %*% V1[5:8])
c(W2[511], h %*% V1[1021:1024])
c(W2[512], h %*% V1[c(1023:1024, 1:2)])
```

@@

c(W3[  1], h %*% V2[1:4])
c(W3[  2], h %*% V2[3:6])
c(W3[  3], h %*% V2[5:8])
c(W3[255], h %*% V2[509:512])
c(W3[256], h %*% V2[c(511:512, 1:2)])
```

@@

First you dilute the W and V values at the top of the pyramid. Then you compute a running dot product of the reversed g vector with the diluted V9 values plus a running dot product of the reversed h vector with the diluted W9 values.

```{r reconstuct-V8}
V9_diluted <- rep(0, 8)
V9_diluted[(1:4)*2-1] <- V9
W9_diluted <- rep(0, 8)
W9_diluted[(1:4)*2-1] <- W9

@@

c(V8[1], rev(g) %*% V9_diluted[c(6, 7, 8, 1)] + rev(h) %*% W9_diluted[c(6, 7, 8, 1)])
c(V8[2], rev(g) %*% V9_diluted[c(7, 8, 1, 2)] + rev(h) %*% W9_diluted[c(7, 8, 1, 2)])
c(V8[3], rev(g) %*% V9_diluted[c(8, 1, 2, 3)] + rev(h) %*% W9_diluted[c(8, 1, 2, 3)])
c(V8[4], rev(g) %*% V9_diluted[c(1, 2, 3, 4)] + rev(h) %*% W9_diluted[c(1, 2, 3, 4)])
c(V8[5], rev(g) %*% V9_diluted[c(2, 3, 4, 5)] + rev(h) %*% W9_diluted[c(2, 3, 4, 5)])
c(V8[6], rev(g) %*% V9_diluted[c(3, 4, 5, 6)] + rev(h) %*% W9_diluted[c(3, 4, 5, 6)])
c(V8[7], rev(g) %*% V9_diluted[c(4, 5, 6, 7)] + rev(h) %*% W9_diluted[c(4, 5, 6, 7)])
c(V8[8], rev(g) %*% V9_diluted[c(5, 6, 7, 8)] + rev(h) %*% W9_diluted[c(5, 6, 7, 8)])
```

To move from V8 to V7, dilute the V8 and W8 vectors and apply a running dot product again.

```{r reconstuct-V7}
V8_diluted <- rep(0, 16)
V8_diluted[(1:8)*2-1] <- V8
W8_diluted <- rep(0, 16)
W8_diluted[(1:8)*2-1] <- W8

wt.7 <- mdwt(ecg, g, L=7)
V7 <- wt.7$y[1:16]

c(V7[ 1], rev(g) %*% V8_diluted[c(14, 15, 16,  1)] + rev(h) %*% W8_diluted[c(14, 15, 16,  1)])
c(V7[ 2], rev(g) %*% V8_diluted[c(15, 16,  1,  2)] + rev(h) %*% W8_diluted[c(15, 16,  1,  2)])
c(V7[ 3], rev(g) %*% V8_diluted[c(16,  1,  2,  3)] + rev(h) %*% W8_diluted[c(16,  1,  2,  3)])
c(V7[ 4], rev(g) %*% V8_diluted[c( 1,  2,  3,  4)] + rev(h) %*% W8_diluted[c( 1,  2,  3,  4)])
c(V7[16], rev(g) %*% V8_diluted[c(13, 14, 15, 16)] + rev(h) %*% W8_diluted[c(12, 13, 14, 15)])
```

Here is the reconstruction of V6

```{r reconstuct-V6}
V7_diluted <- rep(0, 32)
V7_diluted[(1:16)*2-1] <- V7
W7_diluted <- rep(0, 32)
W7_diluted[(1:16)*2-1] <- W7

wt.6 <- mdwt(ecg, g, L=6)
V6 <- wt.6$y[1:32]

c(V6[ 1], rev(g) %*% V7_diluted[c(30, 31, 32,  1)] + rev(h) %*% W7_diluted[c(30, 31, 32,  1)])
c(V6[ 2], rev(g) %*% V7_diluted[c(31, 32,  1,  2)] + rev(h) %*% W7_diluted[c(31, 32,  1,  2)])
c(V6[ 3], rev(g) %*% V7_diluted[c(32,  1,  2,  3)] + rev(h) %*% W7_diluted[c(32,  1,  2,  3)])
c(V6[ 4], rev(g) %*% V7_diluted[c( 1,  2,  3,  4)] + rev(h) %*% W7_diluted[c( 1,  2,  3,  4)])
c(V6[32], rev(g) %*% V7_diluted[c(29, 30, 31, 32)] + rev(h) %*% W7_diluted[c(29, 30, 31, 32)])
```

The process for reconstructing V5, V4, etc. is similar. At the final step, dilute V1 and W1. The running dot product will recreate the original signal.

What exactly do V9, V8, V6, etc. measure? They are coefficients associated with the mother wavelet and reconstruct a signal with high frequency events removed.

```{r graph-v}

V9_coefficients <- rep(0, 2048)
V9_coefficients[1:4] <- V9
V9_signal <- as.vector(midwt(V9_coefficients, d4.rwt$h.0, L=9)$x)

V8_coefficients <- rep(0, 2048)
V8_coefficients[1:8] <- V8
V8_signal <- as.vector(midwt(V8_coefficients, d4.rwt$h.0, L=8)$x)

V7_coefficients <- rep(0, 2048)
V7_coefficients[1:16] <- V7
V7_signal <- as.vector(midwt(V7_coefficients, d4.rwt$h.0, L=7)$x)

V6_coefficients <- rep(0, 2048)
V6_coefficients[1:32] <- V6
V6_signal <- as.vector(midwt(V6_coefficients, d4.rwt$h.0, L=6)$x)

wt.5 <- mdwt(ecg, g, L=5)
V5 <- wt.5$y[1:64] 
V5_coefficients <- rep(0, 2048)
V5_coefficients[1:64] <- V5
V5_signal <- as.vector(midwt(V5_coefficients, d4.rwt$h.0, L=5)$x)

wt.4 <- mdwt(ecg, g, L=4)
V4 <- wt.4$y[1:128] 
V4_coefficients <- rep(0, 2048)
V4_coefficients[1:128] <- V4
V4_signal <- as.vector(midwt(V4_coefficients, d4.rwt$h.0, L=4)$x)

wt.3 <- mdwt(ecg, g, L=3)
V3 <- wt.3$y[1:256] 
V3_coefficients <- rep(0, 2048)
V3_coefficients[1:256] <- V3
V3_signal <- as.vector(midwt(V3_coefficients, d4.rwt$h.0, L=3)$x)

wt.2 <- mdwt(ecg, g, L=2)
V2 <- wt.2$y[1:512] 
V2_coefficients <- rep(0, 2048)
V2_coefficients[1:512] <- V2
V2_signal <- as.vector(midwt(V2_coefficients, d4.rwt$h.0, L=2)$x)

wt.1 <- mdwt(ecg, g, L=1)
V1 <- wt.1$y[1:1024] 
V1_coefficients <- rep(0, 2048)
V1_coefficients[1:1024] <- V1
V1_signal <- as.vector(midwt(V1_coefficients, d4.rwt$h.0, L=1)$x)

{
  graph_without_axis(V9_signal, "V9", limits=range(ecg))
  graph_without_axis(V8_signal, "V8", limits=range(ecg))
  graph_without_axis(V7_signal, "V7", limits=range(ecg))
  graph_without_axis(V6_signal, "V6", limits=range(ecg))
  graph_without_axis(V5_signal, "V5", limits=range(ecg))
  graph_without_axis(V4_signal, "V4", limits=range(ecg))
  graph_without_axis(V3_signal, "V3", limits=range(ecg))
  graph_without_axis(V2_signal, "V2", limits=range(ecg))
  graph_without_axis(V1_signal, "V1", limits=range(ecg))
  graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
}
```

What exactly do W9, W8, W7, etc. measure? They are coefficients associated with the father wavelet and reconstruct a signal with at each frequency level.

```{r graph-w}

W9_coefficients <- rep(0, 2048)
W9_coefficients[5:8] <- W9
W9_signal <- as.vector(midwt(W9_coefficients, d4.rwt$h.0, L=9)$x)

W8_coefficients <- rep(0, 2048)
W8_coefficients[9:16] <- W8
W8_signal <- as.vector(midwt(W8_coefficients, d4.rwt$h.0, L=9)$x)

W7_coefficients <- rep(0, 2048)
W7_coefficients[17:32] <- W7
W7_signal <- as.vector(midwt(W7_coefficients, d4.rwt$h.0, L=9)$x)

W6_coefficients <- rep(0, 2048)
W6_coefficients[33:64] <- W6
W6_signal <- as.vector(midwt(W6_coefficients, d4.rwt$h.0, L=9)$x)

W5_coefficients <- rep(0, 2048)
W5_coefficients[65:128] <- W5
W5_signal <- as.vector(midwt(W5_coefficients, d4.rwt$h.0, L=9)$x)

W4_coefficients <- rep(0, 2048)
W4_coefficients[129:256] <- W4
W4_signal <- as.vector(midwt(W4_coefficients, d4.rwt$h.0, L=9)$x)

W3_coefficients <- rep(0, 2048)
W3_coefficients[257:512] <- W3
W3_signal <- as.vector(midwt(W3_coefficients, d4.rwt$h.0, L=9)$x)

W2_coefficients <- rep(0, 2048)
W2_coefficients[513:1024] <- W2
W2_signal <- as.vector(midwt(W2_coefficients, d4.rwt$h.0, L=9)$x)

W1_coefficients <- rep(0, 2048)
W1_coefficients[1025:2048] <- W1
W1_signal <- as.vector(midwt(W1_coefficients, d4.rwt$h.0, L=9)$x)

{
  graph_without_axis(W9_signal, "W9", limits=range(ecg))
  graph_without_axis(W8_signal, "W8", limits=range(ecg))
  graph_without_axis(W7_signal, "W7", limits=range(ecg))
  graph_without_axis(W6_signal, "W6", limits=range(ecg))
  graph_without_axis(W5_signal, "W5", limits=range(ecg))
  graph_without_axis(W4_signal, "W4", limits=range(ecg))
  graph_without_axis(W3_signal, "W3", limits=range(ecg))
  graph_without_axis(W2_signal, "W2", limits=range(ecg))
  graph_without_axis(W1_signal, "W1", limits=range(ecg))
  graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
}
```
